---
title: "Identify the distribution for each model"
author: "Morgan Gray"
format:
  # pdf:
  #   toc: true
  #   toc-title: Contents
  #   number-sections: true
  #   colorlinks: true
  html:
    toc: true
    toc-title: Contents
    number-sections: TRUE
editor: visual
execute:
  echo: false
  warning: false
---

```{r setup}
# Load libraries, functions, workflows -----
rm(list = ls())
#
# library(tidyverse) ## To manipulate data frames
library(here) ## To manage directories
library(readr)
library(dplyr)
library(forcats)
library(vcd)
library(fitdistrplus)
library(MASS)
library(pscl)
library(car)
library(kableExtra)
library(sessioninfo)

#
source(here("R/functions/fxn_utilities.R"))
# Read the richness and abundance data tables
source(here("R/functions/fxn_load_rich_abun.R"))


```

```{r functions}
fxn_kable <- function(df){
  
  df  %>%
    knitr::kable() %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"), 
                  full_width = TRUE,  
                  position = "left", 
                  fixed_thead = TRUE)
}

fxn_histogram <- function(data, data_name) {
  hist(data$value, breaks = 15, col = "gray", main = paste("Histogram of", data_name), xlab = "Value", ylab = "Count")
  
  if ("value_std" %in% names(data)) {
    hist(data$value_std, breaks = 25, col = "gray", main = paste("Histogram of Standardized", data_name), xlab = "Standardized Value", ylab = "Count")
  }
  
  if ("value_log" %in% names(data)) {
    hist(data$value_log, breaks = 25, col = "gray", main = paste("Histogram of Log-transformed", data_name), xlab = "Log Value", ylab = "Count")
  }
  
  if ("value_sqrt" %in% names(data)) {
    hist(data$value_sqrt, breaks = 25, col = "gray", main = paste("Histogram of Square Root-transformed", data_name), xlab = "Sqrt Value", ylab = "Count")
  }
}

fxn_gof_test_rich <- function(data, data_name) {
  gf_pois <- vcd::goodfit(data$value, type = "poisson")
  pois_first_line <- summary(gf_pois)[1, ]
  plot(gf_pois, main = "") 
  
  gf_nbin <- vcd::goodfit(data$value, type = "nbinomial")
  nbin_first_line <- summary(gf_nbin)[1, ]
  plot(gf_nbin, main = "") 
}

fxn_gof_test_abun <- function(data, data_name) {
  print(paste("Normal Distribution Fit for", data_name)) 
  print(shapiro.test(data$value))
  fitdistrplus::fitdist(data$value, dist = "norm") %>%
    plot(histo = TRUE, demp = TRUE)
  cat("\n")
  
  print(paste("Normal Distribution Fit for Standardized", data_name)) 
  print(shapiro.test(data$value_std))
  fitdistrplus::fitdist(data$value_std, dist = "norm") %>%
    plot(histo = TRUE, demp = TRUE)
  cat("\n")
  
  print(paste("Normal Distribution Fit for Log-transformed", data_name))
  print(shapiro.test(data$value_log))
  fitdistrplus::fitdist(data$value_log, dist = "norm") %>%
    plot(histo = TRUE, demp = TRUE)
  cat("\n")
  
  print(paste("Normal Distribution Fit for Square Root-transformed", data_name)) 
  print(shapiro.test(data$value_sqrt))
  fitdistrplus::fitdist(data$value_sqrt, dist = "norm") %>%
    plot(histo = TRUE, demp = TRUE)
}

fxn_vuong_test <- function(data, data_name) {
  if (length(unique(data$treatment)) > 1) {
    tryCatch({
      nb <- MASS::glm.nb(value ~ treatment, data = data)
      p <- glm(value ~ treatment, data = data, family = "poisson")
      print(paste("Vuong test: Negative Binomial vs. Poisson for", data_name))
      print(pscl::vuong(nb, p))
      cat("\n\n")
      
      if (min(data$value) == 0) {
        z <- pscl::zeroinfl(value ~ treatment, data = data)
        print(paste("Vuong test: Poisson vs. Zero-Inflated for", data_name))
        print(pscl::vuong(p, z))
        cat("\n\n")
        print(paste("Vuong test: Negative Binomial vs. Zero-Inflated for", data_name))
        print(pscl::vuong(nb, z))
      }
    }, error = function(e) {
      message(paste("Error in Vuong tests for", data_name, ":", e$message))
    })
  } else {
    message("Skipping Vuong tests as no treatment variable is present.")
  }
}

fxn_qq_plot <- function(data, data_name) {
  data$value_1 <- data$value + 1
  
  poisson <- MASS::fitdistr(data$value_1, "Poisson")
  car::qqp(data$value_1, "pois", lambda = poisson$estimate, main = paste("Poisson QQ plot for", data_name))
  
  if (data_name == "nonnative richness") {
    car::qqp(data$value_1, "norm", main = paste("Normal QQ plot for", data_name))
    car::qqp(data$value_1, "lnorm", main = paste("Lognormal QQ plot for", data_name))
    
    gamma <- MASS::fitdistr(data$value_1, "gamma")
    car::qqp(data$value_1, "gamma", shape = gamma$estimate[[1]], rate = gamma$estimate[[2]], main = paste("Gamma QQ plot for", data_name))
  }
}



```

# Introduction

The objective of this analysis was to determine the most suitable probability distribution for each of six generalized linear mixed models (GLMMs) fit to plant observation data. Two response variable types were considered: species richness (count data) and percent cover abundance (continuous data). For each response variable, models were fit for three plant groups: native species, native forb species (a subset of native species), and non-native species.

::: callout-note
The subsequent model fitting, selection, and performance assessments are not described here.
:::

# Methods

I conducted a systematic analysis combining visual inspection, statistical tests, and model comparisons to identify the most suitable probability distributions for count and continuous data. Modular functions were developed to handle each assessment component, including visualization generation and statistical testing. This approach provided multiple lines of evidence for selecting the most appropriate distribution for subsequent analyses.

The evaluation framework included:

-   Visual inspection via histograms and quantile-quantile (QQ) plots

-   Distribution fitting tests specific to data type:

    -   Count data: goodness-of-fit tests and pairwise Vuong tests

    -   Continuous data: Shapiro-Wilks test

The influence of data transformations (standardized, log, and square root) was also explored. Log transformation was applied to positively skewed data or data with non-negative values, while square root transformation was used to reduce skewness in count-like or non-negative data. A constant of 1 was added to all values prior to log transformation for distributions requiring strictly positive values.

## Histograms

The analysis for each data type began with visual inspection of histograms generated from the raw data. These visualizations revealed key distributional features, including central tendency, spread, symmetry, and potential outliers.

For count data, histogram shapes indicated potential alignment with common distributions like Poisson (right-skewed with single peak) or negative binomial (right-skewed with longer tail). For continuous data, histograms were generated for raw, standardized, log-transformed, and square-root transformed values.

## Goodness-of-fit

For richness (count) data, discrete goodness-of-fit tests were applied to evaluate Poisson and negative binomial distributions. P-values from these tests were interpreted as follows:

-   *p* \> 0.05: The data were consistent with the distribution (failure to reject the null hypothesis).
-   *p* \< 0.05: The data significantly deviated from the distribution (rejection of the null hypothesis).

For abundance (continuous) data, the Shapiro-Wilk test was used to evaluate normality of both raw and transformed data, using the same significance thresholds.

### Vuong tests

For count data, Vuong tests were used to directly compare the fit of competing models (i.e., Poisson vs. negative binomial). The test produces a z-statistic and p-value, where:

-   Positive z-statistics (p \< 0.05): First model provides better fit

-   Negative z-statistics (p \< 0.05): Second model provides better fit

-   p \> 0.05: Models are statistically indistinguishable

### Quantile-quantile plots

QQ plots compared observed data against theoretical distributions. While not providing discrete test statistics, these visualizations supported statistical test results and offered insights when data failed to fit common distributions. I evaluated plot patterns focusing on:

-   Overall adherence to the diagonal reference line

-   Nature of deviations (random vs. systematic)

-   Patterns at distribution extremes

Points falling along the diagonal reference line indicated good agreement between observed and theoretical distributions. Deviations from the line, particularly systematic patterns, suggested departures from the theoretical distribution. In ecological count data, deviations at the extremes, often due to rare species (excess zeros) or highly abundant counts, were examined for their implications for model selection (e.g., zero-inflated models or negative binomial models).

For count data, QQ plots were reviewed for Poisson and negative binomial distributions. When initial tests were inconclusive (e.g., for non-native counts), QQ plots for alternative distributions (normal, log-normal, gamma) were examined based on histogram shapes.

In ecological count data, deviations are often observed at the extremes, particularly due to rare species (excess zeros) or highly abundant counts. These patterns can aid in model selection. For example, consistent deviations at low values might suggest the need for a zero-inflated model, while heavy tails might favor a negative binomial over a Poisson distribution.

### Additional plots for continuous data

For abundance (continuous) data, QQ plots were generated for normal distributions fit to raw, standardized, square-root transformed, and log-transformed values. In addition to QQ plots, density plots (overlaid on histograms), cumulative distribution function (CDF) plots, and probability-probability (PP) plots were used to visualize the comparison between empirical and theoretical distributions.

# Results

Final recommended distribution (and transformation, when relevant) for each of the six models provided in table. Detailed results for each assessment in subsequent sections.

Native and native forbs for each metric were identical. Negative binomial distribution recommended for native richness because it was a count; normal distribution ok for non-native richness. Recommended distribution for abundance was normal for all species subsets.

| Metric    | Species     | Distribution      | Transformation |
|-----------|-------------|-------------------|----------------|
| Richness  | Native      | Negative binomial | None           |
| Richness  | Native forb | Negative binomial | None           |
| Richness  | Non-native  | Normal            | Log            |
| Abundance | Native      | Normal            | Log            |
| Abundance | Native forb | Normal            | Log            |
| Abundance | Non-native  | Normal            | Square root    |

: *Recommended distribution and transformation for each of the six models*

<br>

## Richness

Preview the first 10 rows of the data table for richness (rich) to see the column names and formats.

::: callout-tip
Scroll to the right to see more columns.
:::

```{r richness}
rich %>%
  dplyr::slice(1:10) %>%
  fxn_kable()
```

### Native species

::: {.callout-note icon="false"}
#### Recommended distribution

Negative binomial
:::

#### **Histograms**

The histograms indicated the raw data followed a Poisson distribution, which was expected for count data.

-   Raw values: Skewed left with tail

-   Log-transformed values: More normal but with outliers to the far left (negative values)

-   Square root-transformed values: Kindof normal with some gaps in the lower range

```{r }
fxn_histogram(rich_nat, "native richness")

```

#### **Goodness-of-fit tests**

The goodness-of-fit test results suggested a negative binomial distribution was a good fit for the data.

-   Poisson distribution: poor (p \< 0.05)

-   Negative binomial distribution: ok (p \> 0.05)

```{r }
fxn_gof_test_rich(rich_nat, "native richness")

```

#### **Vuong tests**

The Vuong test results supported a negative binomial distribution for the data.

-   Negative binomial vs. Poisson: Negative binomial better (positive z, p \< 0.05)

-   Poisson vs. zero-inflated: No difference (p \> 0.05)

-   Negative binomial vs. zero-inflated: Negative binomial better (positive z, p \< 0.05)

```{r }
fxn_vuong_test(rich_nat, "native richness")

```

#### **QQ plot(s)**

The QQ plot supported a negative binomial distribution for the data. The points approximately followed the line, although there was some zero-inflation (lower values above the line) and skew above the line with upper values.

```{r }
fxn_qq_plot(rich_nat, "native richness")

```

<br>

### Native forb species

::: {.callout-note icon="false"}
#### Recommended distribution

Negative binomial
:::

The results for native forb species richness were similar to those for all native species. This was not surprising given the forb data was a large component of the native species data.

#### **Histograms**

The histograms indicated the raw data followed a Poisson distribution, which was expected for count data.

-   Raw values: Skewed left with tail

-   Log-transformed values: More normal but with outliers to the far left (negative values)

-   Square root-transformed values: Kindof normal with some gaps in the lower range

```{r }
fxn_histogram(rich_frb, "native forb richness")

```

#### **Goodness-of-fit tests**

The goodness-of-fit test results suggested a negative binomial distribution was a good fit for the data.

-   Poisson distribution: poor (p \< 0.05)

-   Negative binomial distribution: good (p \> 0.05); better than for native species

```{r }
fxn_gof_test_rich(rich_frb, "native forb richness")

```

#### **Vuong tests**

The Vuong test results supported a negative binomial distribution for the data.

-   Negative binomial vs. Poisson: Negative binomial better (positive z, p \< 0.05)

-   Poisson vs. zero-inflated: No difference (p \> 0.05)

-   Negative binomial vs. zero-inflated: Negative binomial better (positive z, p \< 0.05)

```{r }
fxn_vuong_test(rich_frb, "native forb richness")

```

#### **QQ plot(s)**

The QQ plot supported a negative binomial distribution for the data. The points approximately followed the line, although there was some zero-inflation (lower values above the line).

```{r }
fxn_qq_plot(rich_frb, "native forb richness")

```

<br>

### Non-native species

::: {.callout-note icon="false"}
#### Recommended distribution

Normal (with log-transformed values)
:::

#### **Histograms**

The histogram for the square-root transformed data appeared the most normal of those considered.

-   Raw values: Slightly skewed to the left

-   Standardized: Slightly skewed to the left

-   Log-transformed values: Skewed to the right

-   Square root-transformed values: Kindof normal with some gaps in the lower range

```{r }
fxn_histogram(rich_non, "non-native richness")

```

#### **Goodness-of-fit tests**

The goodness-of-fit test results suggested a Poisson distribution was an ok fit for the raw data.

-   Poisson distribution: ok (p \> 0.05)

-   Negative binomial distribution: poor (p \< 0.05)

```{r }
fxn_gof_test_rich(rich_non, "non-native richness")

```

#### **Vuong tests**

The Vuong test results supported a Poisson distribution for the data.

-   Negative binomial vs. Poisson: Poisson better (negative z, p \< 0.05)

```{r }
fxn_vuong_test(rich_non, "non-native richness")

```

#### **QQ plot(s)**

The QQ plot for the Poisson distribution was not a strong fit for the data. The points approximately followed the line, although never within the bounds; lower values were below the line and upper values were above line.

```{r }
fxn_qq_plot(rich_non, "non-native richness")

```

<br>

------------------------------------------------------------------------

## Abundance

Preview the first 10 rows of the data table for abundance (abun) to see the column names and formats.

::: callout-tip
Scroll to the right to see more columns.
:::

```{r abundance}
abun %>%
  dplyr::slice(1:10) %>%
  fxn_kable()
```

### Native species

::: {.callout-note icon="false"}
#### Recommended distribution

Normal (with log-transformed values)
:::

#### **Histograms**

The histograms indicated the log-transformed data most closely resembled a normal distribution.

-   Raw values: Skewed left, long tail to right

-   Standardized values: Skewed left between -1 to 0, long tail to right

-   Log-transformed values: Normal distribution with outlier to far left (negative)

-   Square root-transformed values: Main peak at 3, additional peak(s) between 7-10

```{r }
fxn_histogram(abun_nat, "native abundance")

```

#### Goodness-of-fit

A comparison of theoretical and empirical distributions showed the best fit for log-transformed values.

-   Raw values: Different central tendency and spread for empirical and theoretical density plot; curvature away from the reference line in the QQ plot, PP plot, and CDF. The Shapiro-Wilk normality test results for indicated the data significantly deviated from a normal distribution.

-   Standardized values: Same as for raw values

-   Log-transformed values: Similar central tendency and spread for density curves; points relatively consistent with reference line for QQ plot, PP plot, and CDF. The Shapiro-Wilk normality test results for indicated the data significantly deviated from a normal distribution.

-   Square root-transformed values: Different central tendency, spread, and symmetry for empirical and theoretical density plot; modest curvature away from the reference line in the QQ plot (especially at low values), PP plot, and CDF. While the W statistic suggests a reasonable degree of normality, the very small p-value suggests that the deviation from normality is statistically significant.

```{r}
fxn_gof_test_abun(abun_nat, "native abundance")
```

<br>

### Native forb species

::: {.callout-note icon="false"}
#### Recommended distribution

Normal (with log-transformed values)
:::

The results for native forb species abundance were similar to those for all native species. This was not surprising given the forb data was a large component of the native species data.

#### Histograms

The histograms indicated the log-transformed data most closely resembled a normal distribution.

-   Raw values: Skewed left, long tail to right

-   Standardized values: Skewed left between -1 to 0, long tail to right

-   Log-transformed values: Normal distribution with outlier to far left (negative)

-   Square root-transformed values: Main peak at 3, broad plateau (no clear peak) of elevated counts between 4-10

```{r }
fxn_histogram(abun_frb, "native forb abundance")

```

#### Goodness-of-fit

A comparison of theoretical and empirical distributions showed the best fit for log-transformed values. The square root-transformed values appeared to behave well upon visual inspection. I opted for the log-transformation in part to be consistent with the approach used for native species and also based on model selection results. Specifically, I compared the performance of models fit to the log-transformed and square-root transformed values (results not shown here) and found the log transformation to be a better approach.

-   Raw values: Different central tendency and spread for empirical and theoretical density plot; curvature away from the reference line in the QQ plot, PP plot, and CDF. The Shapiro-Wilk normality test results for indicated the data significantly deviated from a normal distribution.

-   Standardized values: Same as for raw values

-   Log-transformed values: Similar central tendency and spread for density curves; points relatively consistent with reference line for CDF; low values fall below the reference line for QQ plot and PP plot. The Shapiro-Wilk normality test results for indicated the data significantly deviated from a normal distribution.

-   Square root-transformed values: Different central tendency, spread, and symmetry for empirical and theoretical density plot; modest curvature away from the reference line in the QQ plot, PP plot, and CDF. While the W statistic suggests a reasonable degree of normality, the very small p-value suggests that the deviation from normality is statistically significant.

```{r}
fxn_gof_test_abun(abun_frb, "native forb abundance")
```

<br>

### Non-native species

::: {.callout-note icon="false"}
#### Recommended distribution

Normal (with square root-transformed values)
:::

#### Histograms

The histograms indicated the square root-transformed data most closely resembled a normal distribution.

-   Raw values: Skewed left, tail to right

-   Standardized values: Skewed left between -1 to 0, tail to right

-   Log-transformed values: Skewed right with tail to left

-   Square root-transformed values: Kindof normal

```{r }
fxn_histogram(abun_non, "non-native abundance")

```

#### Goodness-of-fit

A comparison of theoretical and empirical distributions showed the best fit for square root-transformed values.

-   Raw values: Different central tendency and symmetry for empirical and theoretical density plot; curvature away from the reference line for upper values in the QQ plot; points relatively consistent with reference line for PP plot, and CDF. While the W statistic suggests a high degree of normality, the significant p-value indicates a deviation from normality.

-   Standardized values: Same as for raw values

-   Log-transformed values: Different central tendency and symmetry for empirical and theoretical density plot; modest curvature away from the reference line in the QQ plot, PP plot, and CDF. While the W statistic suggests a reasonable degree of normality, the very small p-value suggests that the deviation from normality is statistically significant.

-   Square root-transformed values: Similar central tendency and spread for density curves (empirical still isn't completely symmetrical); points relatively consistent with reference line for the QQ plot, PP plot, and CDF. Even though the W statistic is close to 1 (suggesting near-normality), the p-value (0.0059) provides evidence that the data significantly deviates from a normal distribution. But this was the best fit among the options.

```{r}
fxn_gof_test_abun(abun_non, "non-native abundance")
```

------------------------------------------------------------------------

# Session info

```{r}

## Packages and environment
sessioninfo::session_info()
```
