---
title: "Model selection for richness"
author: "Morgan Gray"
format: 
  html:
    toc: true
    toc-title: Contents
    
editor: visual
execute:
  echo: false
  warning: false
callout-icon: false
number-sections: TRUE
---

```{r setup}
# Load libraries, functions, workflows -----
rm(list = ls())
#
# library(tidyverse) ## To manipulate data frames
library(here)      # To manage directories
library(readr)
library(dplyr)     # For data manipulation
library(forcats)
library(sessioninfo)

source(here("R/functions/fxn_utilities.R"))
source(here("R/functions/fxn_load_rich_abun.R"))

set.seed(912)



```

```{r functions}
# Load required packages ----
# library(lme4)      # For mixed-effects models
# library(MuMIn)     # For model selection
# library(ggplot2)   # For visualization
# library(DHARMa)    # Model diagnostics and residual checks 
# library(performance) 
# library(broom.mixed)
# library(patchwork) # For combining plots
# library(ggplotify)
library(glmmTMB)

# Function for tables ----
fxn_kable <- function(df){
  
  df  %>%
    knitr::kable() %>%
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"), 
                  full_width = TRUE,  
                  position = "left", 
                  fixed_thead = TRUE)
}
```

# Introduction

This document details the model selection process for richness models developed as part of a larger project analyzing plant observations. Three generalized linear mixed models (GLMMs) were fit to predict plant richness (species count, a positive integer) for different plant groups. These groups included all native plants (nat), native forbs (frb, a subset of native plants), and non-native plants (non). For each group, a best-fit model was selected from a candidate set based on model performance. This document describes the methods used for model selection, summarizes the model diagnostics and decision making process, and presents the final selected models.

# Methods

Text

# Results

The final models are presented in the table below, with detailed model selection information provided in subsequent sections. The model for non-native species used log-transformed richness, a normal distribution, and the same fixed and random effects structure as the abundance models (fixed effects for grazing treatment and survey year, and a random intercept and slope for treatment within plot name).

The final models for native and native forb species were identical, employing untransformed values, a negative binomial distribution, and the same fixed and random effects structure as the non-native model, with the addition of a fixed effect for plot type (PER, WAN).

| Species | Response | Formula | Distribution |
|--------------|--------------|------------------------------|--------------|
| Native | Richness | grazing + survey year + plot type + (1 + grazing \| plot) | Negative binomial |
| Native forb | Richness | grazing + survey year + plot type + (1 + grazing \| plot) | Negative binomial |
| Non-native | log(Richness) | grazing + survey year + (1 + grazing \| plot) | Normal |

: *Final richness models*

------------------------------------------------------------------------

## Native species

::: {.callout-note icon="false"}
#### Final model for native species richness

log(richness) \~ grazing + survey year + plot type + (1 + grazing \| plot)
:::

The log-transformed response with fixed effects for grazing treatment and survey year, and a random intercept and a random slope for treatment within plot name emerged as the best model for native species richness.

### Initial model assessment

The native species richness was previously identified as best fitting a normal distribution with log-transformed values (see *Identify the distribution for each model*). In contrast, the AIC comparison from this initial model assessment showed the standardized values had a better fit than log-transformed values.

```{r}
# Fit the initial models
models_init_rich_nat <- fxn_initial_model(rich_nat)

# View the summary table
models_init_rich_nat$summary_table %>%
  fxn_kable()
```

However, model diagnostics for standardized values didn't look good. Visual inspection of QQ plot showed marked deviation from reference; and the tests for deviation, outliers, and within-group deviations from uniformity were significant.

```{r}
# Review diagnostics for model with lowest AIC
fxn_model_review(models_init_rich_nat$best_model$model, rich_nat)
```

The next best set of transformations ranked by AIC were log-transformed values. These model diagnostics looked better than those for the standardized values: less deviation shown in the QQ plot, and no significant outliers. Test for within-group deviations from uniformity was significant.

```{r}
# Identify the model with the lowest AIC for each transformation
models_init_rich_nat$summary_table %>%
  group_by(response) %>%
  slice(1) %>%
  arrange(aic) %>%
  ungroup()

```

These model diagnostics looked better than those for the standardized values: less deviation shown in the QQ plot, and no significant outliers. Test for within-group deviations from uniformity was significant.

```{r}
# lmer_2: lmer(richness ~ treatment + (1 + treatment | plot_name))
model_init_rich_nat_log <- lmer(
  value_log ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_nat", REML = FALSE
)
fxn_model_review(model_init_rich_nat_log, rich_nat)

```

Because the model diagnostics for the log-transformed values were better than those for the standardized values and because previous work supported log-transformation for native species richness, I elected to use the model for log-tranformed values with the lowest AIC. The log-transformed response with treatment as a fixed effect, and a random intercept and a random slope for treatment within plot name emerged as the best initial model.

#### Best initial model

log(richness) \~ treatment + (1 + treatment \| plot_name)

<br>

### Fixed effects selection

The AIC results for the fixed effects models showed the model with a fixed effect for f_one_year to be the best fit for the data. The models that included f_break and f_new were within 3 AIC of the minimum. These models all related to sampling timing (i.e., 1-year gaps between surveys, a survey after a prolonged break, or the first survey year).

All models with one additional fixed effect were within 6 AIC of the minimum.

```{r}
# Define the best initial model
best_model_init_rich_nat <- lmer(
  value_log ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_nat", REML = FALSE
)

# Fit the fixed effects models
models_fixed_rich_nat <- fxn_fixed_effects(best_model_init_rich_nat)

# View the summary table
fixed_summary_table <- models_fixed_rich_nat$summary_table %>%
  as_tibble() %>%
  dplyr::select(model_name, terms_count, aic) %>%
  dplyr::mutate(delta = round(aic - min(aic), 1))

fixed_summary_table %>%
  fxn_kable()

# View the best model 
best_fixed_model <- models_fixed_rich_nat$best_model$formula

# Capture the output of print()
# output <- capture.output(print(models_fixed_rich_nat$best_model$formula)) 

```

The model diagnostics for the four top-performing models showed deviation consistent with the initial model (not worse, at least) and no significant outliers. However, tests for within-group deviations from uniformity were significant for treatment and for the fixed effect.

::: {.callout-tip icon="false"}
#### Note

The diagnostic plots for excluded fixed effect models fit to native species richness are included in the Appendix for reference.
:::

```{r}

model_fixed_rich_nat_5 <- update(best_model_init_rich_nat, . ~ . + f_year)

fxn_model_review(model_fixed_rich_nat_5, rich_nat)

```

The addition of sampling year as a fixed effect emerged as the best model after fixed effects selection based on AIC values and diagnostic plots. After reviewing the model diagnostics for all the fixed effects selection models, I selected the model with a fixed term for sampling year (f_year). Although this model did not have the lowest AIC, it was within the pre-defined threshold of 6 AIC units for consideration (delta = 4.3). My decision was influenced by the lack of within-group deviations from uniformity. Specifically, the effect of year was uniform with homogenous variance. The inclusion of a fixed effect for year also improved the uniformity for grazed plots.

#### Best model after fixed effects selection

log(richness) \~ grazing + survey year + (1 + grazing \| plot)

<br>

### Random effects selection

The addition of random effects resulted in singular fit for all combinations. No random effects were added to the model.

```{r}
# Define the best fixed effects model
best_model_fixed_rich_nat <- update(best_model_init_rich_nat, . ~ . + f_year)

models_random_rich_nat <- fxn_random_effects(best_model_fixed_rich_nat)

models_random_rich_nat %>%
  fxn_kable()
```

<br>

### Final model selection

The final selected model included log-transformed richness as the response variable, with fixed effects for grazing treatment and survey year, and a random intercept and a random slope for treatment within plot name. This model satisfied assumptions of uniform residual distribution and proper dispersion, showed no significant outliers, and maintained convergence without singularity issues.

#### Final model for native species richness

log(richness) \~ grazing + survey year + (1 + grazing \| plot)

<br>

------------------------------------------------------------------------

## Native forb species

#### Final model for native forb species richness

### Initial model assessment

Text

```{r}
# Fit the initial models
models_init_rich_frb <- fxn_initial_model(rich_frb)

# View the summary table
models_init_rich_frb$summary_table %>%
  fxn_kable()
```

Text

```{r}
# Review diagnostics for model with lowest AIC
fxn_model_review(models_init_rich_frb$best_model$model, rich_frb)
```

Text

```{r}
# Identify the model with the lowest AIC for each transformation
models_init_rich_frb$summary_table %>%
  group_by(response) %>%
  slice(1) %>%
  arrange(aic) %>%
  ungroup()

# lmer_2: lmer(richness ~ treatment + (1 + treatment | plot_name))
model_init_rich_frb_log <- lmer(
  value_log ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_frb", REML = FALSE
)
fxn_model_review(model_init_rich_frb_log, rich_frb)

```

Text

```{r}

```

Text

#### Best initial model

<br>

### Fixed effects selection

Text

```{r}
# Define the best initial model
best_model_init_rich_frb <- lmer(
  value_log ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_frb", REML = FALSE
)

# Fit the fixed effects models
models_fixed_rich_frb <- fxn_fixed_effects(best_model_init_rich_frb)

# View the summary table
fixed_summary_table <- models_fixed_rich_frb$summary_table %>%
  as_tibble() %>%
  dplyr::select(model_name, terms_count, aic) %>%
  dplyr::mutate(delta = round(aic - min(aic), 1))

fixed_summary_table %>%
  fxn_kable()

# View the best model 
best_fixed_model <- models_fixed_rich_frb$best_model$formula


```

Text

```{r}

model_fixed_rich_frb_5 <- update(best_model_init_rich_frb, . ~ . + f_year)

fxn_model_review(model_fixed_rich_frb_5, rich_frb)
```

Text

#### Best model after fixed effects selection

<br>

### Random effects selection

Text

```{r}
# Define the best fixed effects model
best_model_fixed_rich_frb <- update(best_model_init_rich_frb, . ~ . + f_year)

models_random_rich_frb <- fxn_random_effects(best_model_fixed_rich_frb)

models_random_rich_frb %>%
  fxn_kable()
```

Text

```{r}

```

Text

<br>

### Final model selection

Text

<br>

------------------------------------------------------------------------

## Non-native species

### Initial model assessment

Text

```{r}
# Fit the initial models
models_init_rich_non <- fxn_initial_model(rich_non)

# View the summary table
models_init_rich_non$summary_table %>%
  fxn_kable()
```

Text

```{r}
# Review diagnostics for model with lowest AIC
fxn_model_review(models_init_rich_non$best_model$model, rich_non)
```

Text

```{r}
# Identify the model with the lowest AIC for each transformation
models_init_rich_non$summary_table %>%
  group_by(response) %>%
  slice(1) %>%
  arrange(aic) %>%
  ungroup()

# lmer_2: lmer(richness ~ treatment + (1 + treatment | plot_name))
model_init_rich_non_log <- lmer(
  value_log ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_non", REML = FALSE
)
fxn_model_review(model_init_rich_non_log, rich_non)

```

Text

```{r}
# lmer_2: lmer(richness ~ treatment + (1 + treatment | plot_name))
model_init_rich_non_sqrt <- lmer(
  value_sqrt ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_non", REML = FALSE
)
fxn_model_review(model_init_rich_non_sqrt, rich_non)

```

Text

#### Best initial model

<br>

### Fixed effects selection

Text

```{r}
# Define the best initial model
best_model_init_rich_non <- lmer(
  value_log ~ treatment + (1 + treatment | plot_name),
  data = rich, subset = met_sub == "rich_non", REML = FALSE
)

# Fit the fixed effects models
models_fixed_rich_non <- fxn_fixed_effects(best_model_init_rich_non)

# View the summary table
fixed_summary_table <- models_fixed_rich_non$summary_table %>%
  as_tibble() %>%
  dplyr::select(model_name, terms_count, aic) %>%
  dplyr::mutate(delta = round(aic - min(aic), 1))

fixed_summary_table %>%
  fxn_kable()

# View the best model 
best_fixed_model <- models_fixed_rich_non$best_model$formula


```

Text

```{r}

model_fixed_rich_non_5 <- update(best_model_init_rich_non, . ~ . + f_year)

fxn_model_review(model_fixed_rich_non_5, rich_non)
```

Text

#### Best model after fixed effects selection

<br>

### Random effects selection

Text

```{r}
# Define the best fixed effects model
best_model_fixed_rich_non <- update(best_model_init_rich_non, . ~ . + f_year)

models_random_rich_non <- fxn_random_effects(best_model_fixed_rich_non)

models_random_rich_non %>%
  fxn_kable()
```

Text

```{r}

```

Text

<br>

### Final model selection

Text

#### Final model for non-native species richness

Text
