---
title: "Model selection for abundance"
date: now
date-format: "YYYY-MM-DD"
format: 
  pdf:
    toc: true
    toc-title: Contents
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
      - heightrounded
    
editor: visual
execute:
  echo: false
  warning: false
callout-icon: false
number-sections: FALSE
latex-output-dir: ../output

mainfont: "Gill Sans Nova Light" 
sansfont: "Gill Sans Nova Light"
fontsize: 11pt
code-overflow: wrap
---

```{r setup}
# Load libraries, functions, workflows -----
# rm(list = ls())
#
library(DHARMa)    # Model diagnostics and residual checks 
library(dplyr)     # For data manipulation
library(forcats)
library(ggplot2)   # For visualization
library(here)      # To manage directories
library(lme4)      # For mixed-effects models
library(MuMIn)     # For model selection
library(performance) 
library(readr)
library(sessioninfo)

source(here::here("R/1_setup.R"))
source(here("R/2_model_selection_abundance.R"))
# source(here("R/functions/fxn_lmer_initial_model.R"))
# source(here("R/functions/fxn_lmer_fixed_effects.R"))
# source(here("R/functions/fxn_lmer_random_effects.R"))
# source(here("R/functions/fxn_lmer_model_review.R"))

set.seed(912)

```

# Introduction

This document details the model selection process for abundance models developed as part of a larger project analyzing plant observations. Three generalized linear mixed models (GLMMs) were fit to predict plant abundance (percent plot cover, a continuous value from 0 to 1) for different plant groups. These groups included all native plants (nat), native forbs (frb, a subset of native plants), and non-native plants (non). For each group, a best-fit model was selected from a candidate set based on model performance. This document describes the methods used for model selection, summarizes the model diagnostics and decision making process, and presents the final selected models.

# Methods

I used a systematic, three-stage approach to identify the optimal model structure for analyzing plant abundance data: initial model assessment, fixed effects selection, and random effects selection. Modular functions were developed to automate portions of the model selection process, which was completed for each of the three groups (nat, frb, non). The following sections describe this process.

All models were fit using restricted maximum likelihood (REML) via the `lmer` function from the `lme4` package (Bates et al. 2015). Model comparison was primarily based on Akaike's Information Criterion corrected for small sample size (AICc), calculated using the `MuMIn` package (Barton 2022). Models within a delta AICc of 6 were considered for further evaluation. Formal tests for uniform residual distribution (`testUniformity`), dispersion (`testDispersion`), and outlier detection (`testOutliers`) were performed using the `DHARMa` package (Hartig 2022). Diagnostic plots, including quantile-quantile plots (`DHARMa::plotQQunif`) and residual plots (`DHARMa::plotResiduals`), were visually inspected. Model singularity was assessed using the `performance` package (Lüdecke et al., 2021), specifically the `check_singularity` function. The `janitor` package (Firshman, 2023) was used for data cleaning and table formatting.

## Initial model assessment

For each group (nat, frb, non), I first assessed the influence of response variable transformation on model fit by fitting several models. These included intercept-only and simple linear models using untransformed abundance as null models:

-   `lm_null`: An intercept-only linear model using untransformed abundance `lm(abundance ~ 1)`
-   `lm`: A linear model with untransformed abundance and a fixed effect for treatment: `lm(abundance ~ treatment)`

Linear mixed-effects models were then fit using standardized, log-transformed, and square-root transformed abundance. For each transformation, four model structures were compared:

-   `lmer_null_1`: A mixed-effects model with a random intercept for plot name (intercept-only): `lmer(transformed abundance ~ 1 + (1 | plot_name))`
-   `lmer_null_2`: A mixed-effects model with a random intercept and a random slope for treatment within plot name (without treatment as a fixed effect): `lmer(transformed abundance ~ 1 + (1 + treatment | plot_name))`
-   `lmer_1`: A mixed-effects model with treatment as a fixed effect and a random intercept for plot name: `lmer(transformed abundance ~ treatment + (1 | plot_name))`
-   `lmer_2`: A mixed-effects model with treatment as a fixed effect, and a random intercept and a random slope for treatment within plot name: `lmer(transformed abundance ~ treatment + (1 + treatment | plot_name))`

AICc values were compared to identify the most promising transformation. QQ plots were used to assess residual normality for each transformation. I also considered the results from a separate exploratory analysis conducted to determine the most suitable probability distribution for each species group (see *Identify the distribution for each model*).

## Fixed effects selection

Based on the best-performing transformation from the initial model assessment, I fit a series of models with different combinations of fixed effects. Starting with the best performing initial model, I iteratively added or removed fixed effects: `plot_type`, `f_year`, `f_break`, `f_new`, `f_one_yr`, and `f_two_yr`.

An initial exploration was conducted using automated model dredging (via the `dredge` function from the `MuMIn` package), followed by manual model comparisons. The `dredge` function explores all possible predictor combinations and ranks them by AICc. While dredge results provided a comprehensive overview of potential variable influences and informed subsequent model selection, the dredge results themselves were not used to identify the final best-fit model.

I developed functions to streamline the fixed effect selection process. First, a function individually added each potential fixed effect to the best initial model and compared AICc values to identify the best single-term model. Next, a separate function added a second term to the best single-term model and compared AICc values to identify the best two-term model. Models with a delta AICc less than 6 were retained. Residual plots were examined for each model to assess model assumptions.

## Random effects selection

After identifying a set of best-performing fixed effects models, I explored the inclusion of additional random effects. I tested the addition of random intercepts for other variables individually. I also tested the addition of random slopes for treatment within each predictor variable.

Model comparisons were again based on AICc, retaining models within a delta AICc of 6. Singularity checks were performed using the `lme4::check_singularity` function to ensure model convergence.

| Abbreviation | Model specification |
|-----------------------|------------------------------------------------|
| plot_type | best_fixed_model, . \~ . + (1 \| plot_type) |
| f_year | best_fixed_model, . \~ . + (1 \| f_year) |
| f_year_grazer | best_fixed_model, . \~ . + (1 \| f_year/grazer) |
| grazer | best_fixed_model, . \~ . + (1 \| grazer) |
| f_break | best_fixed_model, . \~ . + (1 \| f_break) |
| f_one_yr | best_fixed_model, . \~ . + (1 \| f_one_yr) |
| f_two_yr | best_fixed_model, . \~ . + (1 \| f_two_yr) |
| f_new | best_fixed_model, . \~ . + (1 \| f_new) |
| f_year_plot_type | best_fixed_model, . \~ . + (1 + f_year \| plot_type) |
| treatment_f_year | best_fixed_model, . \~ . + (1 + treatment \| f_year) |
| treatment_f_year_grazer | best_fixed_model, . \~ . + (1 + treatment \| f_year/grazer) |
| treatment_grazer | best_fixed_model, . \~ . + (1 + treatment \| grazer) |
| treatment_f_break | best_fixed_model, . \~ . + (1 + treatment \| f_break) |
| treatment_f_one_yr | best_fixed_model, . \~ . + (1 + treatment \| f_one_yr) |
| treatment_f_two_yr | best_fixed_model, . \~ . + (1 + treatment \| f_two_yr) |
| treatment_f_new | best_fixed_model, . \~ . + (1 + treatment \| f_new) |

: *Abbreviations used for random effect model selection*

## Final model selection

The final model for each group was selected based on AICc, diagnostic plot review, and ecological interpretability. I chose the model with the lowest AICc that met model assumptions and had a biologically plausible structure. Final model performance was assessed by examining residual plots, testing for uniform residual distribution using `testUniformity` (from `DHARMa`), and assessing for outliers. The rationale for each predictor's inclusion in the final model was documented.

# Results

The final models are presented in the table below, with detailed model selection information provided in subsequent sections. While all three models used a normal distribution and the same fixed and random effects structure (fixed effects for grazing treatment and survey year, and a random intercept and slope for treatment within plot name), the transformation of the abundance data differed slightly. The final models for native and native forb species were identical, using log-transformed abundance. The non-native model used a square root transformation of abundance.

| Species | Response | Formula | Distribution |
|-----------|-------------|------------------------------------|------------|
| Native | log(Abundance) | grazing + survey year + plot type + (grazing \| plot) | Normal |
| Native forb | log(Abundance) | grazing + survey year + plot type + (grazing \| plot) | Normal |
| Non-native | sqrt(Abundance) | grazing + survey year + 2-yr interval + (1 \| plot) | Normal |

: *Final abundance models*

<br>

## Native species

::: {.callout-note icon="false"}
#### Final model for native species abundance

log(Abundance) \~ grazing + survey year + plot type + (grazing \| plot)
:::

The log-transformed response with fixed effects for grazing treatment and survey year, and a random intercept and a random slope for treatment within plot name emerged as the best model for native species abundance.

### Initial model assessment

The native species abundance was previously identified as best fitting a normal distribution with log-transformed values (see *Identify the distribution for each model*). In contrast, the AIC comparison from this initial model assessment showed the standardized values had a better fit than log-transformed values.

```{r}
# View the summary table
models_init_abun_nat$summary_table %>%
  fxn_kable()
```

However, model diagnostics for standardized values didn't look good. Visual inspection of QQ plot showed marked deviation from reference; and the tests for deviation, outliers, and within-group deviations from uniformity were significant.

```{r}
# Review diagnostics for model with lowest AIC
fxn_lmer_model_review(models_init_abun_nat$best_model$model, abun_nat)
```

The next best set of transformations ranked by AIC were log-transformed values. These model diagnostics looked better than those for the standardized values: less deviation shown in the QQ plot, and no significant outliers. Test for within-group deviations from uniformity was significant.

```{r}
# Identify the model with the lowest AIC for each transformation
lowest_aic_init_abun_nat
```

These model diagnostics looked better than those for the standardized values: less deviation shown in the QQ plot, and no significant outliers. Test for within-group deviations from uniformity was significant.

```{r}
# lmer_2: lmer(abundance ~ treatment + (1 + treatment | plot_name))
model_init_abun_nat_log 

fxn_lmer_model_review(model_init_abun_nat_log, abun_nat)

```

Because the model diagnostics for the log-transformed values were better than those for the standardized values and because previous work supported log-transformation for native species abundance, I elected to use the model for log-tranformed values with the lowest AIC. The log-transformed response with treatment as a fixed effect, and a random intercept and a random slope for treatment within plot name emerged as the best initial model.

::: {.callout-note icon="false"}
#### Best initial model

log(Abundance) \~ treatment + (1 + treatment \| plot_name)
:::

<br>

### Fixed effects selection

The AIC results for the fixed effects models showed the model with a fixed effect for f_one_year to be the best fit for the data. The models that included f_break and f_new were within 3 AIC of the minimum. These models all related to sampling timing (i.e., 1-year gaps between surveys, a survey after a prolonged break, or the first survey year).

All models with one additional fixed effect were within 6 AIC of the minimum.

```{r}
fixed_summary_table_abun_nat %>%
  fxn_kable()

# View the best model 
best_fixed_model_abun_nat

```

The model diagnostics for the four top-performing models showed deviation consistent with the initial model (not worse, at least) and no significant outliers. However, tests for within-group deviations from uniformity were significant for treatment and for the fixed effect.

```{r}

model_fixed_abun_nat_5 

fxn_lmer_model_review(model_fixed_abun_nat_5, abun_nat)

```

The addition of sampling year as a fixed effect emerged as the best model after fixed effects selection based on AIC values and diagnostic plots. After reviewing the model diagnostics for all the fixed effects selection models, I selected the model with a fixed term for sampling year (f_year). Although this model did not have the lowest AIC, it was within the pre-defined threshold of 6 AIC units for consideration (delta = 4.3). My decision was influenced by the lack of within-group deviations from uniformity. Specifically, the effect of year was uniform with homogenous variance. The inclusion of a fixed effect for year also improved the uniformity for grazed plots.

::: callout-note
#### Best model after fixed effects selection

log(Abundance) \~ grazing + survey year + (grazing \| plot)
:::

<br>

### Random effects selection

The addition of random effects resulted in singular fit for all combinations. No random effects were added to the model.

```{r}
models_random_abun_nat %>%
  fxn_kable()
```

<br>

### Final model selection

The final selected model included log-transformed abundance as the response variable, with fixed effects for grazing treatment and survey year, and a random intercept and a random slope for treatment within plot name. This model satisfied assumptions of uniform residual distribution and proper dispersion, showed no significant outliers, and maintained convergence without singularity issues.

::: {.callout-note icon="false"}
#### Final model for native species abundance

log(Abundance) \~ grazing + survey year + (grazing \| plot)
:::

<br>

------------------------------------------------------------------------

## Native forb species

::: {.callout-note icon="false"}
#### Final model for native forb species abundance

log(Abundance) \~ grazing + survey year + plot type + (grazing \| plot)

(same model as for native species overall)
:::

### Initial model assessment

Fit the initial set of candidate models and review the AICc values.

```{r}
# View the summary table
models_init_abun_frb$summary_table %>%
  fxn_kable()
```

Review the diagnostics for the model with the lowest AICc

```{r}
# Review diagnostics for model with lowest AIC
fxn_lmer_model_review(models_init_abun_frb$best_model$model, abun_frb)
```

Review the diagnostics for alternate models

```{r}
# Identify the model with the lowest AIC for each transformation
lowest_aic_init_abun_frb

# lmer_2: lmer(abundance ~ treatment + (1 + treatment | plot_name))
fxn_lmer_model_review(model_init_abun_frb_log, abun_frb)

```

Text

::: callout-note
#### Best initial model

log(Abundance) \~ treatment + (1 + treatment \| plot_name)
:::

<br>

### Fixed effects selection

Fit the fixed effects one at a time and review the AICc values

```{r}
# View the summary table
fixed_summary_table_abun_frb %>%
  fxn_kable()

best_fixed_model_abun_frb
```

Review the model diagnostics

```{r}

model_fixed_abun_frb_5

fxn_lmer_model_review(model_fixed_abun_frb_5, abun_frb)
```

Text

::: callout-note
#### Best model after fixed effects selection
:::

<br>

### Random effects selection

Review the AICc values with additional random effects

```{r}
models_random_abun_frb %>%
  fxn_kable()
```

Define the final model

<br>

------------------------------------------------------------------------

## Non-native species

### Initial model assessment

Text

```{r}
models_init_abun_non$summary_table %>%
  fxn_kable()
```

Review the model diagnostics

```{r}
# Review diagnostics for model with lowest AIC
fxn_lmer_model_review(models_init_abun_non$best_model$model, abun_non)
```

Identify the model with the lowest AICc for each transformation

```{r}
# Identify the model with the lowest AIC for each transformation
lowest_aic_init_abun_non


```

Review diagnostics for alternate models

```{r}
fxn_lmer_model_review(model_init_abun_non_sqrt, abun_non)

```

Text

::: callout-note
#### Best initial model
:::

<br>

### Fixed effects selection

View the fixed effects summary table

```{r}
fixed_summary_table_abun_non %>%
  fxn_kable()

# View the best model 
best_fixed_model_abun_non 


```

Review the model diagnostics

```{r}
fxn_lmer_model_review(model_fixed_abun_non_5, abun_non)
```

Text

::: callout-note
#### Best model after fixed effects selection
:::

<br>

### Random effects selection

Text

```{r}
# Define the best fixed effects model
models_random_abun_non %>%
  fxn_kable()
```

Text

<br>

### Final model selection

Text

::: {.callout-note icon="false"}
#### Final model for non-native species abundance
:::

Text

------------------------------------------------------------------------

# References

-   Barton, K. (2022). *MuMIn: Multi-Model Inference*. R package version 1.46.0.

-   Bates, D., Mächler, M., Bolker, B. M., & Walker, S. C. (2015). Fitting Linear Mixed-Effects Models Using lme4. *Journal of Statistical Software*, *67*(1), 1–48.

-   Firshman, T. (2023). *janitor: Simple Data Frame Cleaning Functions*. R package version 2.2.0.

-   Hartig, F. (2022). *DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models*. R package version 0.4.6.

-   Lüdecke, D., Ben-Shachar, M., Patil, I., Waggoner, P., & Makowski, D. (2021). performance: Assessing and Comparing Statistical Models. *Journal of Open Source Software*, *6*(60), 3139.

------------------------------------------------------------------------

------------------------------------------------------------------------

# Session info

```{r}
## Packages and environment
sessioninfo::session_info()
```
