---
title: "Variable summaries"
author: "Morgan Gray"
date: now
date-format: "YYYY-MM-DD"
format: 
  pdf:
    toc: true
    toc-title: Contents
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
      - heightrounded
    
editor: visual
execute:
  echo: false
  warning: false
callout-icon: false
number-sections: FALSE
latex-output-dir: ../output

mainfont: "Gill Sans Nova Light" 
sansfont: "Gill Sans Nova Light"
fontsize: 11pt
code-overflow: wrap
---

```{r setup}
# Run setup scripts
source(here::here("R/1_setup.R"))

# Create functions to summarize the metrics by variable group
# source(here::here("R/functions/fxn_summarize_models.R"))

# Create function for pretty tables 
# source(here::here("R/functions/fxn_kable.R"))

# Create input data subsets 
# abun <- rich_abun$abundance  
# abun_nat <- abun$abun_nat 
# abun_frb <- abun$abun_frb
# abun_non <- abun$abun_non
# 
# rich <- rich_abun$richness
# rich_nat <- rich$rich_nat
# rich_frb <- rich$rich_frb
# rich_non <- rich$rich_non


```

# Introduction

This document summarizes the data subsets used to analyze plant community observations. Six generalized linear mixed models (GLMMs) were fit to examine two aspects of plant communities: species richness (count data) and percent cover abundance (continuous proportions from 0-1). For each response variable, three plant groups were analyzed: all native species, native forbs (a subset of native species), and non-native species. For each combination of response variable and plant group, the best-performing model from a candidate set of models was selected.

Explanatory variables used in the model:

Treatment: Grazed vs ungrazed

Survey year

Plot type

Plot pair

Grazing interval: 1-year, 2-year, new

# Abundance

```{r}
# mod_abun_nat <- lme4::lmer(
#   value_log ~ treatment + f_year +  plot_type + (1 + treatment | plot_name), 
#   data = abun_nat, REML = FALSE)

```

{{< pagebreak >}}

# Environment

**Session info**

```{r}
sessionInfo()
```
