---
title: "Model summaries"
author: "Morgan Gray"
date: now
date-format: "YYYY-MM-DD"
format: 
  pdf:
    toc: true
    toc-title: Contents
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
      - heightrounded
    
editor: visual
execute:
  echo: false
  warning: false
callout-icon: false
number-sections: FALSE
latex-output-dir: ../output

mainfont: "Gill Sans Nova Light" 
sansfont: "Gill Sans Nova Light"
fontsize: 11pt
code-overflow: wrap
---

```{r setup}
# Run setup scripts
source(here::here("R/1_setup.R"))

# Create functions to summarize the models
source(here::here("R/functions/fxn_summarize_models.R"))

# Create function for pretty tables 
source(here::here("R/functions/fxn_kable.R"))

# Create functions to tidy tables 
fxn_tidy_table <- function(index_table){
  
  index_table %>%
    dplyr::mutate(estimate = round(bt_estimate, 2), 
                  conf_low = round(bt_conf_low, 2),
                  conf_high = round(bt_conf_high, 2), 
                  statistic = round(statistic, 2)) %>%
    dplyr::select(dplyr::any_of(c(
      # "subset", 
      "treatment",
      "term",
      "contrast",
      "estimate", 
      "conf_low", 
      "conf_high",
      "p_value", 
      "statistic")))
  
}

# Create input data subsets 
abun <- rich_abun$abundance  
abun_nat <- abun$abun_nat 
abun_frb <- abun$abun_frb
abun_non <- abun$abun_non

rich <- rich_abun$richness
rich_nat <- rich$rich_nat
rich_frb <- rich$rich_frb
rich_non <- rich$rich_non


```

<br>

# Introduction

This document summarizes statistical models developed to analyze plant community observations. Six generalized linear mixed models (GLMMs) were fit to examine two aspects of plant communities: species richness (count data) and percent cover abundance (continuous proportions from 0-1). For each response variable, three plant groups were analyzed: all native species, native forbs (a subset of native species), and non-native species. For each combination of response variable and plant group, the best-performing model from a candidate set of models was selected.

The formulation, model summary, marginal means, and contrasts are provided in this document for each of the six models.

{{< pagebreak >}}

# Abundance

## Native species

```{r echo = TRUE}
mod_abun_nat <- lme4::lmer(
  value_log ~ treatment + f_year +  plot_type + (1 + treatment | plot_name), 
  data = abun_nat, REML = FALSE)
```

**Model summary table**

```{r}
summary(mod_abun_nat)
```

{{< pagebreak >}}

**Marginal means**

```{r}
mm_abun_nat <- fxn_summarize_marginal_means("mod_abun_nat", lookup_tables)  

fxn_tidy_table(mm_abun_nat) %>%
  fxn_kable()

```

**Contrasts**

```{r}
contrasts_abun_nat <- fxn_summarize_contrasts("mod_abun_nat", lookup_tables)  

fxn_tidy_table(contrasts_abun_nat) %>%
  fxn_kable()

```

<br>

{{< pagebreak >}}

## Native forb species

```{r echo = TRUE}
mod_abun_frb <- lme4::lmer(
  value_log ~ treatment + f_year +  plot_type + (1 + treatment | plot_name), 
  data = abun_frb,  REML = FALSE)
```

**Model summary table**

```{r}
summary(mod_abun_frb)
```

{{< pagebreak >}}

**Marginal means**

```{r}
mm_abun_frb <- fxn_summarize_marginal_means("mod_abun_frb", lookup_tables)  

mm_abun_frb_tbl <- fxn_tidy_table(mm_abun_frb) 

mm_abun_frb_tbl %>%
  fxn_kable()
```

**Contrasts**

```{r}
contrasts_abun_frb <- fxn_summarize_contrasts("mod_abun_frb", lookup_tables)  

fxn_tidy_table(contrasts_abun_frb) %>%
  fxn_kable()
```

{{< pagebreak >}}

## Non-native species

```{r echo = TRUE}
mod_abun_non <- lme4::lmer(
  value_sqrt ~ treatment + f_year + f_two_yr + (1 | plot_name), 
  data = abun_non, REML = FALSE)
```

**Model summary table**

```{r}
summary(mod_abun_non)
```

{{< pagebreak >}}

**Marginal means**

```{r}
mm_abun_non <- fxn_summarize_marginal_means("mod_abun_non", lookup_tables)  

fxn_tidy_table(mm_abun_non) %>%
  fxn_kable()
```

**Contrasts**

```{r}
contrasts_abun_non <- fxn_summarize_contrasts("mod_abun_non", lookup_tables)  

fxn_tidy_table(contrasts_abun_non) %>%
  fxn_kable()

```

{{< pagebreak >}}

# Richness

## Native species

```{r echo = TRUE}
mod_rich_nat <- glmmTMB::glmmTMB(
  value ~ 
    treatment + 
    f_year + 
    (1 | plot_type) + 
    (1 + treatment | plot_name),
  data = rich_nat, 
  family = nbinom2,
  control = glmmTMBControl(optimizer=optim, 
                           optArgs=list(method="BFGS")))

```

**Model summary table**

```{r}
summary(mod_rich_nat)
```

{{< pagebreak >}}

**Marginal means**

```{r}
mm_rich_nat <- fxn_summarize_marginal_means("mod_rich_nat", lookup_tables)  
# 
fxn_tidy_table(mm_rich_nat) %>%
  fxn_kable()
```

**Contrasts**

```{r}
contrasts_rich_nat <- fxn_summarize_contrasts("mod_rich_nat", lookup_tables)  

fxn_tidy_table(contrasts_rich_nat) %>%
  fxn_kable()
```

{{< pagebreak >}}

## Native forb species

```{r echo = TRUE}
mod_rich_frb <- glmmTMB::glmmTMB(
  value ~ 
    treatment + 
    f_year + 
    (1 | plot_type) + 
    (1 + treatment | plot_name),
  data = rich_frb, 
  family = nbinom2,
  control = glmmTMBControl(optimizer=optim, 
                           optArgs=list(method="BFGS")))

```

**Model summary table**

```{r}
summary(mod_rich_frb)
```

{{< pagebreak >}}

**Marginal means**

```{r}
mm_rich_frb <- fxn_summarize_marginal_means("mod_rich_frb", lookup_tables)  
# 
fxn_tidy_table(mm_rich_frb) %>%
  fxn_kable()
```

**Contrasts**

```{r}
contrasts_rich_frb <- fxn_summarize_contrasts("mod_rich_frb", lookup_tables)  

fxn_tidy_table(contrasts_rich_frb) %>%
  fxn_kable()

```

{{< pagebreak >}}

## Non-native species

```{r echo = TRUE}
mod_rich_non <- lme4::lmer(
  value_sqrt ~ 
    treatment + 
    f_year + 
    (1 + treatment | plot_name), 
  data = rich_non, 
  REML = FALSE)
```

**Model summary table**

```{r}
summary(mod_rich_non)
```

{{< pagebreak >}}

**Marginal means**

```{r}
mm_rich_non <- fxn_summarize_marginal_means("mod_rich_non", lookup_tables)  

fxn_tidy_table(mm_rich_non) %>%
  fxn_kable()
```

**Contrasts**

```{r}
contrasts_rich_non <- fxn_summarize_contrasts("mod_rich_non", lookup_tables)  

fxn_tidy_table(contrasts_rich_non) %>%
  fxn_kable()

```

{{< pagebreak >}}

# Environment

You must run the scripts in fxn_summarize_models() with marginaleffects_0.5.0. If you use a later version you could (and probably will) get different results.

Get marginaleffects_0.5.0 from cran by running these 2 lines:

```{r echo=TRUE}
# Uncomment (remove the #) code to run

# packageurl <- "https://cran.r-project.org/src/contrib/Archive/marginaleffects/marginaleffects_0.5.0.tar.gz"
# install.packages(packageurl, repos=NULL, type="source")
```

Then confirm the package version:

```{r echo=TRUE}
# Uncomment (remove the #) code to run

# sessionInfo()
```

------------------------------------------------------------------------

**Session info**

```{r}
sessionInfo()
```
